% Tutorial 4: Terra/Luna Collapse - ECOM215
\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage[left=1in,top=1in,right=1in,bottom=1in,headheight=3ex,headsep=3ex]{geometry}
\usepackage{graphicx}
\usepackage{float}
\usepackage{booktabs}
\usepackage{enumitem}

\usepackage[sc]{mathpazo}
\linespread{1.05}
\usepackage[T1]{fontenc}

\usepackage{fancyhdr,lastpage}
\pagestyle{fancy}
\usepackage{hyperref}
\usepackage{lastpage}

\usepackage{array, xcolor}
\usepackage{colortbl}
\definecolor{clemsonorange}{HTML}{EA6A20}
\definecolor{ImperialBlue}{RGB}{0, 62, 116}
\definecolor{AccentGreen}{RGB}{46, 139, 87}
\hypersetup{colorlinks,breaklinks,linkcolor=clemsonorange,urlcolor=clemsonorange,anchorcolor=clemsonorange,citecolor=black}

\lhead{}
\chead{}
\rhead{\footnotesize ECOM215: Tutorial 4 --- Terra/Luna Collapse}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\title{Tutorial 4: Terra/Luna Collapse\\[0.3em] \Large Anatomy of a Death Spiral\\[1em] \normalsize ECOM215: Blockchain Economics and Digital Assets}
\author{Week 5 $\vert$ Decentralised Finance (DeFi)}
\date{Semester B, 2025/2026}

\begin{document}

\maketitle

\vspace{1em}
\begin{center}
\fbox{\parbox{0.9\textwidth}{\centering\textbf{CASE BRIEF FOR STUDENTS}\\[0.3em] Please read before the tutorial. Estimated reading time: 15 minutes.}}
\end{center}

\section*{Background}

In early May 2022, the Terra blockchain was one of the most celebrated projects in cryptocurrency. Its native stablecoin, UST (TerraUSD), had grown to become the third-largest stablecoin with a market capitalisation of approximately \$18 billion. The associated token, LUNA, had a market cap of around \$40 billion, making it a top-10 cryptocurrency.

One week later, both were worthless. Approximately \$60 billion in value was destroyed in what became the largest collapse in cryptocurrency history.

This case examines what Terra was, how it worked, why it failed, and what lessons it offers for understanding DeFi risks.

\section*{How Terra Worked}

Unlike fiat-backed stablecoins (USDC, USDT) that hold dollar reserves, UST was an \textbf{algorithmic stablecoin}. It maintained its \$1 peg through a mechanism linking it to LUNA:

\medskip
\noindent\textbf{The Core Mechanism:}
\begin{itemize}[leftmargin=*]
\item To \textbf{mint} 1 UST: Burn \$1 worth of LUNA
\item To \textbf{redeem} 1 UST: Burn the UST, receive \$1 worth of LUNA
\end{itemize}

\medskip
\noindent\textbf{How the peg was supposed to work:}
\begin{itemize}[leftmargin=*]
\item If UST trades above \$1: Arbitrageurs mint new UST (by burning LUNA), sell UST on the market, pocket the difference. This increases UST supply, pushing price back to \$1.
\item If UST trades below \$1: Arbitrageurs buy cheap UST on the market, redeem for \$1 worth of LUNA, sell the LUNA. This decreases UST supply, pushing price back to \$1.
\end{itemize}

The system had no external collateral. UST's value depended entirely on the ability to redeem it for LUNA---and LUNA's value depended largely on demand for UST.

\section*{The Anchor Protocol: 20\% Yields}

A critical part of Terra's ecosystem was \textbf{Anchor Protocol}, a lending platform that offered approximately \textbf{20\% APY} on UST deposits.

\medskip
\noindent\textbf{Why this mattered:}
\begin{itemize}[leftmargin=*]
\item 20\% yield on a ``stablecoin'' was extraordinarily attractive
\item At its peak, over 70\% of all UST was deposited in Anchor
\item This created massive demand for UST, supporting LUNA's price
\end{itemize}

\medskip
\noindent\textbf{Where did the yield come from?}
\begin{itemize}[leftmargin=*]
\item Borrowers paid interest (but borrowing demand was insufficient)
\item Staking rewards from collateral
\item \textbf{Subsidies from Terra's reserves}---the yield was not sustainable
\end{itemize}

The 20\% rate was essentially a customer acquisition cost, paid from a ``yield reserve'' that was being depleted. By early 2022, the reserve was shrinking and Anchor announced plans to move to a variable rate.

\section*{The Collapse: May 7--13, 2022}


\noindent\textbf{Saturday, May 7:}
\begin{itemize}[leftmargin=*]
\item Large withdrawals from Anchor (approximately \$2 billion over the weekend)
\item UST begins trading slightly below \$1 (around \$0.98)
\item Unclear whether this was an intentional attack or organic selling
\end{itemize}

\medskip
\noindent\textbf{Sunday--Monday, May 8--9:}
\begin{itemize}[leftmargin=*]
\item UST drops to \$0.90, triggering panic
\item Holders rush to redeem UST for LUNA
\item Each redemption mints new LUNA, increasing LUNA supply
\item LUNA price falls from \$80 to \$60
\end{itemize}

\medskip
\noindent\textbf{Tuesday--Wednesday, May 10--11:}
\begin{itemize}[leftmargin=*]
\item Death spiral accelerates
\item UST drops to \$0.30
\item LUNA supply explodes: from 350 million to over 6 \textit{trillion} tokens
\item LUNA price collapses: \$60 $\rightarrow$ \$1 $\rightarrow$ \$0.01 $\rightarrow$ \$0.0001
\end{itemize}

\medskip
\noindent\textbf{Friday, May 13:}
\begin{itemize}[leftmargin=*]
\item Terra blockchain halted twice
\item UST trading at \$0.10--\$0.15
\item LUNA effectively worthless
\item Combined losses: approximately \$60 billion
\end{itemize}

\section*{The Death Spiral Mechanism}

The collapse illustrated a \textbf{reflexive death spiral}:

\begin{enumerate}[leftmargin=*]
\item UST loses peg slightly
\item Holders redeem UST for LUNA (minting new LUNA)
\item Increased LUNA supply pushes LUNA price down
\item Now \$1 of LUNA is worth less, so more LUNA must be minted per UST redeemed
\item LUNA price falls further
\item Confidence in UST collapses (if LUNA is worthless, UST redemptions are worthless)
\item Everyone tries to exit simultaneously
\item System cannot absorb the selling pressure
\end{enumerate}

The fundamental problem: \textbf{circular backing}. UST was backed by LUNA, but LUNA's value depended on UST demand. When confidence broke, both collapsed together.

\section*{Warning Signs That Were Ignored}

\begin{itemize}[leftmargin=*]
\item \textbf{Unsustainable yields}: 20\% APY with no clear source of returns
\item \textbf{Concentration risk}: 70\%+ of UST in a single protocol (Anchor)
\item \textbf{No external collateral}: Nothing outside the system to absorb shocks
\item \textbf{Previous stress}: UST had briefly lost its peg in May 2021
\item \textbf{Academic warnings}: Researchers had identified the death spiral risk
\item \textbf{Size mismatch}: UST market cap (\$18B) approaching LUNA market cap (\$40B)---less ``buffer''
\end{itemize}

\section*{The Aftermath}

\textbf{Immediate consequences:}
\begin{itemize}[leftmargin=*]
\item Tens of thousands of investors lost substantial savings
\item Some retail investors had put life savings into Anchor for the 20\% yield
\item Contagion spread: Crypto hedge funds (Three Arrows Capital) and lenders (Celsius, Voyager) collapsed in subsequent weeks
\end{itemize}

\medskip
\noindent\textbf{Legal and regulatory:}
\begin{itemize}[leftmargin=*]
\item Do Kwon (Terra founder) charged with fraud by US SEC and South Korean authorities
\item Arrested in Montenegro (March 2023) while travelling on false documents
\item Extradition proceedings ongoing
\item Accelerated regulatory focus on stablecoins globally
\end{itemize}

\medskip
\noindent\textbf{Industry response:}
\begin{itemize}[leftmargin=*]
\item Algorithmic stablecoins largely discredited
\item Greater scrutiny of yield sources (``Where does the yield come from?'')
\item MiCA regulation (EU) explicitly addresses algorithmic stablecoin risks
\end{itemize}

\section*{Key Facts Summary}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Item} & \textbf{Detail} \\
\midrule
Date of collapse & May 7--13, 2022 \\
UST market cap (peak) & $\sim$\$18 billion \\
LUNA market cap (peak) & $\sim$\$40 billion \\
Total value destroyed & $\sim$\$60 billion \\
LUNA supply change & 350 million $\rightarrow$ 6+ trillion \\
LUNA price change & \$80 $\rightarrow$ \$0.0001 \\
Anchor yield & $\sim$20\% APY \\
\bottomrule
\end{tabular}
\end{center}

\section*{Questions to Consider}

\begin{enumerate}[leftmargin=*]
\item Why did so many people invest in a system with such obvious risks? What role did the 20\% yield play?

\item Could the collapse have been prevented once it started? What interventions might have worked?

\item How is Terra/Luna different from a traditional bank run? How is it similar?

\item Should algorithmic stablecoins be banned, regulated, or allowed to fail?

\item What due diligence should investors perform before depositing funds in DeFi protocols?
\end{enumerate}

\section*{Further Reading (Optional)}

\begin{itemize}[leftmargin=*]
\item ``The Fall of Terra: A Timeline'' --- The Block Research
\item Nansen research report on Terra wallet flows
\item SEC complaint against Terraform Labs (February 2023)
\end{itemize}

\newpage

%==============================================================================
% PART B: TA DISCUSSION GUIDE
%==============================================================================

\section*{Session Timeline}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Time} & \textbf{Activity} \\
\midrule
0:00--0:08 & Context setting: summarise the mechanism and collapse \\
0:08--0:20 & Discussion Question 1: Why did people invest? \\
0:20--0:32 & Discussion Question 2: Could it have been stopped? \\
0:32--0:44 & Discussion Question 3: Bank run parallels \\
0:44--0:52 & Discussion Question 4: Regulatory response \\
0:52--1:00 & Synthesis and key takeaways \\
\bottomrule
\end{tabular}
\end{center}

\section*{Discussion Questions with Guidance}

\subsection*{Question 1: Why did people invest despite the risks?}

\textit{``The circular backing and unsustainable yields were visible to anyone who looked. Why did so many people---including sophisticated investors---put money into Terra?''}

\medskip
\noindent\textbf{Points that may emerge:}

\begin{itemize}[leftmargin=*]
\item \textbf{Yield hunger}: In a low-interest-rate environment, 20\% on a ``stablecoin'' was irresistible
\item \textbf{Social proof}: Major VCs (Jump, Three Arrows) invested; ``smart money'' validated it
\item \textbf{Complexity as camouflage}: The mechanism was complicated enough that many didn't understand the risks
\item \textbf{Short-term thinking}: Many knew it was risky but thought they could exit before collapse
\item \textbf{Normalisation}: UST had existed for years and survived a prior stress test (May 2021)
\item \textbf{Motivated reasoning}: People wanted to believe; confirmation bias filtered out warnings
\end{itemize}

\medskip
\noindent\textbf{Key insight}: High yields without clear sources should always trigger scepticism. ``Where does the yield come from?'' is the fundamental due diligence question in DeFi.

\subsection*{Question 2: Could the collapse have been stopped?}

\textit{``Once UST started losing its peg on May 7, was the collapse inevitable? What interventions might have worked?''}

\medskip
\noindent\underline{Potential interventions:}
\begin{itemize}[leftmargin=*]
\item \textbf{Bitcoin reserves}: Luna Foundation Guard had $\sim$\$3B in BTC reserves; they deployed them but it was insufficient
\item \textbf{Circuit breakers}: Pause redemptions temporarily to prevent panic spiral
\item \textbf{External bailout}: A large institution could have provided liquidity (none did)
\item \textbf{Mint caps}: Limit how much LUNA could be minted per day (but this breaks the peg mechanism)
\end{itemize}

\medskip
\noindent\underline{Why interventions failed or wouldn't work:}
\begin{itemize}[leftmargin=*]
\item \$3B reserves vs \$18B UST market cap---insufficient buffer
\item Pausing redemptions destroys confidence in the peg entirely
\item No ``lender of last resort'' in DeFi (unlike central banks for traditional banks)
\item The mechanism itself was the problem---any fix undermines the core design
\end{itemize}

\medskip
\noindent\textbf{Key insight}: Once a death spiral starts in a system with circular backing, it may be impossible to stop. The design was fundamentally fragile, not just poorly managed.

\subsection*{Question 3: How is this like (and unlike) a bank run?}

\textit{``Traditional banks can also experience runs. How is Terra/Luna similar to a bank run? How is it different?''}

\medskip
\noindent\underline{Similarities:}
\begin{itemize}[leftmargin=*]
\item Confidence-driven: Both collapse when depositors lose faith
\item Self-fulfilling: The act of withdrawing causes further losses
\item Speed matters: Slow withdrawals are manageable; panic is not
\item Contagion: Both can spread to connected institutions
\end{itemize}

\medskip
\noindent\underline{Differences:}
\begin{itemize}[leftmargin=*]
\item \textbf{No deposit insurance}: Banks have FDIC/government backstop; DeFi does not
\item \textbf{No lender of last resort}: Central banks provide emergency liquidity; no equivalent in DeFi
\item \textbf{24/7 markets}: Bank runs historically took days; Terra collapsed in hours
\item \textbf{Transparency}: Everyone could watch reserves drain in real-time (accelerated panic)
\item \textbf{Circular backing}: Banks hold external assets (loans, bonds); Terra's backing was internal
\end{itemize}

\medskip
\noindent\textbf{Key insight}: DeFi has recreated traditional financial risks (runs, leverage, contagion) without the safety infrastructure (insurance, central banks, regulation) that evolved over centuries to manage them.

\subsection*{Question 4: What should regulators do about algorithmic stablecoins?}

\textit{``After Terra, regulators worldwide have proposed rules for stablecoins. Should algorithmic stablecoins be banned? Regulated? Left alone to fail?''}

\medskip
\noindent\underline{Arguments for banning:}
\begin{itemize}[leftmargin=*]
\item Retail investors cannot assess the risks
\item No algorithmic stablecoin has achieved long-term stability at scale
\item Systemic risk if they grow large enough
\item Consumer protection mandate
\end{itemize}

\medskip
\noindent\underline{Arguments for regulation (not banning):}
\begin{itemize}[leftmargin=*]
\item Innovation shouldn't be prohibited, just controlled
\item Require clear disclosures, risk warnings, investor caps
\item Reserve requirements or capital buffers
\item Let sophisticated investors participate with informed consent
\end{itemize}

\medskip
\noindent\underline{Arguments for allowing failure:}
\begin{itemize}[leftmargin=*]
\item Caveat emptor---investors should do due diligence
\item Market discipline: Failures teach lessons
\item Regulation may not prevent fraud and gives false sense of security
\item DeFi is global; national bans just push activity elsewhere
\end{itemize}

\medskip
\noindent\textbf{Key insight}: MiCA (EU) essentially requires stablecoins to have reserves, which effectively prohibits pure algorithmic models. The US approach remains unclear. There's genuine tension between innovation and protection.

\subsection*{Extension Question (if time permits)}

\textit{``Do Kwon is facing fraud charges. But the mechanism was public---anyone could read the smart contracts. Is it fraud if everything was transparent?''}

\medskip
This raises interesting questions about disclosure vs.\ fraud. The SEC alleges misrepresentation about the stability mechanism and sustainability of yields, not just that the system failed.

\vfill
\begin{flushright}
\textit{End of Tutorial 4 Materials}
\end{flushright}

\end{document}